Linear and Differential Cryptanalysis of DES

### 2009-01-30 trunk/talk/draft1.tex
SPN Diff slides

Vulnerability of DES:\\
Differential Cryptanalysis exploits the high probability of certain occurrences of differential patterns in the cipher. In DES, this highly probable differentials occur due to
\begin{itemize}
\item{S-Boxes show a strong tendency to produce certain differential pairs with high probability, rather than being random.}
\item{The differences do not get diffused fast enough.}
\item{Differentials are not affected by the round keys.}
\end{itemize}
\end{frame}

@@

\[ \Delta P \rightarrow \Delta X_1 \rightarrow \Delta X_2 \rightarrow \cdots \rightarrow \Delta X_{k-1} \rightarrow \Delta C \]

@@

\begin{frame}
Basic math concepts behind Differential attack ...

@@

\begin{frame}
Intro to differentials ...

\begin{frame}
Differential characteristic / approximation of SBox ...

\begin{frame}
Probabilities for the approximations ...

\begin{frame}
Analogue of Piling Up lemma ...

\begin{frame}
Combining across SPN rounds ...

\begin{frame}
The algorithm for the attack ...

\begin{frame}
Probabilistic justification ...

>>>>

Vulnerability of DES:\\
\begin{itemize}
\item{S-Boxes show a strong tendency to produce certain differential pairs with high probability, rather than being random.}
\item{The differences do not get diffused fast enough.}
\item{Differentials are not affected by the round keys.}
\end{itemize}
\end{frame}

@@

\[ \Delta P \rightarrow \Delta C^{(1)} \rightarrow \Delta C^{(2)} \rightarrow \cdots \rightarrow \Delta C^{(k-1)} \rightarrow \Delta C \]

@@

\begin{frame}
Differential Cryptanalysis tries to exploit the high probability of certain occurrences of differential characteristics $(\Delta P \rightarrow \Delta C)$ in the cipher. This is a chosen ciphertext attack.

[Differential: $(\Delta P, \Delta C) = (P \oplus P', C \oplus C') $]

We have to analyze the cipher in pieces. The basic idea is
\begin{itemize}
\item{Analyze the difference pattern and their probabilities of the S-Box}
\item{Combine the S-Boxes over the rounds to build the differential characteristic of the cipher}
\item{Choose the characteristic with maximum probability and use that to retrieve subkey bits of the cipher}
\end{itemize}
\end{frame}

\begin{frame}
Analyzing the S-Box

Let us consider the S-Box of the SPN cipher as we constructed it:
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item{$4 \times 4$ S-Box}
\item{Input: $X = [X_1 \: X_2 \: X_3 \: X_4]$}
\item{Output: $Y = [Y_1 \: Y_2 \: Y_3 \: Y_4]$}
\item{Difference pair: $(\Delta X, \Delta Y)$}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\includegraphics[totalheight=0.3\textheight]{spn_sbox_unkeyed.jpg}
\end{figure}\end{column}
\end{columns}

Algorithm
\begin{enumerate}
\item{Fix a certain input difference $\Delta X = 0100$, say}
\item{Iterate through the first input $X = \{0000, ... ,1111 \}$}
\item{This fixes the second input $X' = \{0100, ... ,1011\}$ (ordering does not matter)}
\item{Obtain corresponding output difference $\Delta Y = Y \oplus Y'$}
\item{Iterate through steps 2 to 4 for $\Delta X = \{0000, ... ,1111 \}$} 
\end{enumerate}
\end{frame}

\begin{frame}
Sample Difference Pairs for the S-Box

\begin{figure}
\includegraphics[totalheight=0.6\textheight]{spn_sbox_diffpairs.jpg}
\end{figure}

Distribution of differences

Note that, for $\Delta X = 1011$, $\Delta Y = 0010$ occurs $8$ times, out of the possible $16$ times. So, the difference pair $(1011,0010)$ has a probability of occurrence $\frac{8}{16} = \frac{1}{2}$. So, we can tabulate the complete data for the S-Box to check this probabilities.
\end{frame}

\begin{frame}
Difference Distribution Table

\begin{figure}
\includegraphics[totalheight=0.6\textheight]{spn_sbox_disttable.jpg}
\end{figure}

Note that
\begin{itemize}
\item{In an ideal S-Box, we would like all entries to be $1$}
\item{But this table clearly shows bias towards certain pairs}
\item{Just divide the entries by $2^4 = 16$ to get the probabilities, and exploit the scenario of highest probability of occurrence}
\end{itemize}
\end{frame}

\begin{frame}
Why can not we make an ideal S-Box?

Notice that in the difference table
\begin{itemize}
\item{The $(0,0)$ entry is $16$, just because identical inputs ($\Delta X = 0 \Rightarrow X = X'$) should produce identical outputs ($Y = Y' \Rightarrow \Delta Y = 0$)}
\item{All the entries are even, because $\Delta X$ is the same for both the input pairs $(X,X')$ and $(X',X)$, producing same $\Delta Y$}
\item{Sum of all entries in a row is $2^4 = 16$}
\end{itemize}

As an ideal S-Box expects to give out no information of $\Delta Y$ given $\Delta X$, we wish it could have all entries $1$, i.e,
\[ Pr(\Delta Y | \Delta X) = \frac{1}{16} \quad \forall \Delta X  \]
But, from our discussion about the distribution table, this is infeasible.
\end{frame}

\begin{frame}
What happens for a Keyed S-Box?

Let us consider the S-Box of the SPN cipher, with the round keys this time:
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item{$4 \times 4$-bit S-Box}
\item{Input: $W = [W_1 \: W_2 \: W_3 \: W_4]$}
\item{Round Key: $K = [K_1 \: K_2 \: K_3 \: K_4]$}
\item{Output: $Y = [Y_1 \: Y_2 \: Y_3 \: Y_4]$}
\item{Difference pair: $(\Delta W, \Delta Y)$}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\includegraphics[totalheight=0.3\textheight]{spn_sbox_keyed.jpg}
\end{figure}
\end{column}
\end{columns}

Note that for each pair of inputs $(W,W')$, the actual inputs to the S-Box are $(X,X') = (W \oplus K, W' \oplus K)$. Hence, the input difference for the keyed S-Box is 
\[ \Delta W = W \oplus W' = (X \oplus K) \oplus (X' \oplus K) = X \oplus X' = \Delta X \]
and hence $\Delta W \rightarrow \Delta Y$ corresponds identically to $\Delta X \rightarrow \Delta Y$.

Thus, the keyed S-Box has the same difference distribution table as the unkeyed S-Box.
\end{frame}

\begin{frame}
Constructing a Differential Characteristic

\begin{columns}
\begin{column}{0.6\textwidth}

The diagram beside illustrates the tracing of the non-zero bits of the input difference through the SPN structure.

Input Diff: $\Delta P = 0B00$ \\

Propagating Difference Pairs
\begin{itemize}
\item{$S_{12}$: $1011 \rightarrow 0010$ [$Pr = \frac{8}{16}$]}
\item{$S_{23}$: $0100 \rightarrow 0110$ [$Pr = \frac{6}{16}$]}
\item{$S_{32}$: $0010 \rightarrow 0101$ [$Pr = \frac{6}{16}$]}
\item{$S_{33}$: $0010 \rightarrow 0101$ [$Pr = \frac{6}{16}$]}
\item{$S_{ij}$: $0000 \rightarrow 0000$ [$Pr = \frac{16}{16}$]}
\end{itemize}

Differential Characteristic
\[ 0B00 \rightarrow 0040 \rightarrow 0220 \rightarrow 0606 \]
\end{column}
\begin{column}{0.4\textwidth}
\begin{figure}
\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
Piling up Probabilities

Notation
$\Delta U_i$: Input difference to $i$-th round
$\Delta V_i$: Output difference from $i$-th round
$K_i$: Round Key for $i$-th round
$X_{i,j}$: $j$-th bit of $X_i$

Here, we have the following in the differential characteristic
\begin{itemize}
\item{Round 1: $\Delta P = \Delta U_1 = 0B00 \rightarrow 0040 = \Delta V_1$ [$Pr = \frac{8}{16}$]}
\item{Round 2: $\Delta V_1 = \Delta U_2 = 0040 \rightarrow 0220 = \Delta V_2$ [$Pr = \frac{6}{16}$]}
\item{Round 3: $\Delta V_2 = \Delta U_3 = 0220 \rightarrow 0606 = \Delta V_3$ [$Pr = \left(\frac{6}{16}\right)^2$]}
\item{Round 4: $\Delta V_3 = \Delta U_4 = 0606$}
\end{itemize}

Hence, assuming the S-Boxes to be independent over the rounds, we obtain
\[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]
with probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = \frac{8}{16} \times \frac{6}{16} \times \left(\frac{6}{16}\right)^2 = \frac{27}{1024}$.

Note: The probability of this characteristic $Pr = \frac{27}{2^{10}} \gg \frac{1}{2^{16}}$

Now, lets see if we can exploit this situation to extract the key bits.
\end{frame}

\begin{frame}
Attack Idea
\begin{itemize}
\item{We have a differential characteristic that links plaintext difference $\Delta P$ to input difference $\Delta U_4$ of the 4th round output of SPN and holds with high probability ($\frac{27}{1024}$)}
\item{We can partially undo the last round by guessing the last key $K_5$}
\item{If we guess correctly, the differential characteristic will hold with high probability, close to $\frac{27}{1024}$. If we guess wrong, it will hold with low probability}
\item{But in this case, we don't need to guess the entire key $K_5$ for the last round!}
\item{The differential output only involves 2 S-Boxes, $S_{42}$ and $S_{44}$. Thus we only need to guess $2^8 = 256$ values}
\item{For each value of the guessed \textit{target partial subkey} we can undo the last round and determine the probability of the differential characteristic}
\item{Highest probability indicates likely correct guess}
\end{itemize}

@@

\begin{frame}
\begin{columns}
\begin{column}{0.6\textwidth}
Differential Characteristic flow

Input Diff: $\Delta P = 0B00$

Characteristic
\[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]

Probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = \frac{27}{1024}$

Partial Subkey guess only for
\[ [K_{5,5} ... K_{5,8}, K_{5,13} ... K_{5,16}] \]
\end{column}

\begin{column}{0.4\textwidth}
\begin{figure}
\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
\end{figure}
\end{column}
\end{columns}
